---
title: "NTD"
author: "J.Wente"
date: "1/23/2020"
output: html_document
---
#A basic markdown for exploring the National Transit Database data


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(tidyverse)
library(lubridate)
library(stringi)
library(magrittr)
library(readxl)
```


```{r data prep, include=FALSE}
df<-read_excel("November 2019 Raw Database.xlsx",sheet="MASTER")
upt<-read_excel("November 2019 Raw Database.xlsx",sheet="UPT")
vrm<-read_excel("November 2019 Raw Database.xlsx",sheet="VRM")

pop<- df %>% filter(is.na(Agency)==F) %>% select(UZA,`UZA Name`,`UZA Area SQ Miles`,`Service Area Population`,`Service Area SQ Miles`,`Most Recent Report Year`,Active,`Reporter Type`,Agency,Mode)

upt2<-upt %>%  filter(is.na(Agency) == F) %>% 
          gather(year_mon,upt,-Agency,-Active,-`Reporter Type`,-UZA,
                    -`UZA Name`,-Modes,-TOS,-`5 digit NTD ID`,-`4 digit NTD ID`)  %>% 
                  mutate(Month=substr(year_mon, start = 1, stop = 3) %>% as.factor(),
                     Year=substr(year_mon,start=4,stop =5 ),
                     Month=
                       fct_recode(Month,"1"="JAN","2"="FEB","3"="MAR","4"="APR","5"="MAY",
                                      "6"="JUN","7"="JUL","8"="AUG","9"="SEP","10"="OCT",
                                      "11"="NOV","12"="DEC") %>% as.character(),
                     Date=paste0("01","/",Month,"/","20",Year) %>% dmy()
                     )


vrm2<-vrm %>% filter(is.na(Agency) == F) %>% 
          gather(year_mon,vrm,-Agency,-Active,-`Reporter Type`,-UZA,
                    -`UZA Name`,-Modes,-TOS,-`5 digit NTD ID`,-`4 digit NTD ID`) %>% 
              mutate(Month=substr(year_mon, start = 1, stop = 3) %>% as.factor(),
                  Year=substr(year_mon,start=4,stop =5 ),
                  Month=fct_recode(Month,"1"="JAN","2"="FEB","3"="MAR","4"="APR","5"="MAY",
                                      "6"="JUN","7"="JUL","8"="AUG","9"="SEP","10"="OCT",
                                      "11"="NOV","12"="DEC") %>% as.character(),
                  Date=paste0("01","/",Month,"/","20",Year) %>% dmy()
                     ) 

vrm2$MT<-"Oth"
vrm2$MT[vrm2$Agency=="Metro Transit"]<-"MT"
vrm2$MT[vrm2$Agency=="Metro Mobility"]<-"MB"
vrm2$MT[vrm2$Agency=="Metropolitan Council"]<-"METC"

upt2$MT<-"Oth"
upt2$MT[upt2$Agency=="Metro Transit"]<-"MT"
upt2$MT[upt2$Agency=="Metro Mobility"]<-"MB"
upt2$MT[upt2$Agency=="Metropolitan Council"]<-"METC"



mt<-left_join(upt2,vrm2)

mt$pop.uza<-df$`UZA Population`[match(mt$UZA,df$UZA)]
mt$area<-df$`Service Area SQ Miles`[match(mt$UZA,df$UZA)]
mt$pop.sa<-df$`Service Area Population`[match(mt$UZA,df$UZA)]


```

Modes breakdown:
CR = commuter rail
LR = light rail
MB = bus
DR = Demand Response (like Metro Mobility)
PB = Publico. A transit mode comprised of passenger vans or small buses operating with fixed routes but no fixed schedules in Puerto Rico. Publicos (PB) are a privately owned and operated public transit service which is market oriented and unsubsidized, but regulated through a public service commission, state or local government. 
HR = Heavy Rail 
VP = Van pool 

Main question - against other metro areas, how does Metro Transit compare? 

Vehicle Revenue Miles (VRM)
The miles that vehicles are scheduled to or actually travel while in revenue service. Vehicle revenue miles include: 
  • Layover / recovery time.
exclude: 
    • Deadhead; (miles travlled outside of revenue hours/trips)
    •   Operator training;
    •   Vehicle maintenance testing; and
    •   School bus and charter services.

```{r exploratory plots, echo=FALSE}
#A summarized dataset looking at vrm, upt, 
gg0<-mt %>% filter(Modes != "PB",MT %in% c("MB","METC","MT")) %>% mutate(Year=year(Date)) %>%
  group_by(Year) %>% 
    summarize(pop.sa=mean(pop.sa,na.rm=T),vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T),
              area=mean(area,na.rm=T)) %>%
       mutate(vrm.per.pop=vrm/pop.sa,vrm.per.area=vrm/area) %>% ungroup() %>% 
    ggplot(aes(x=Year,y=(vrm/10^6))) + geom_line() + 
    ggtitle("Annual VRM at Met Transit, MB, and METC") #+ 
    #  scale_y_continuous("Million VRM",labels =                                                            scales::comma,breaks=seq(0,40,1)) 
   # scale_x_continuous(breaks=seq(2002,2019,2))
gg0

gg0a<-mt %>% filter(Modes != "PB",MT %in% c("MB","METC","MT")) %>% 
  mutate(Year=year(Date)) %>%
  group_by(Year) %>% 
    summarize(pop.sa=mean(pop.sa,na.rm=T),vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T),
              area=mean(area,na.rm=T)) %>%
       mutate(vrm.per.pop=vrm/pop.sa,vrm.per.area=vrm/area) %>% ungroup() %>% 
    ggplot(aes(x=Year,y=(upt/10^6))) + geom_line() + 
    ggtitle("Annual UPT at Met Transit, MB, and METC") #+ 
     # scale_y_continuous("Million UPT",labels =                                                           # scales::comma,breaks=seq(0,40,1)) 
   # scale_x_continuous(breaks=seq(2002,2019,2))
gg0a

 
  
  


gg1.df<-mt %>% filter(Modes != "PB") %>% group_by(Date,MT,`Reporter Type`,Modes) %>% 
    summarize(pop.sa=mean(pop.sa,na.rm=T),vrm=mean(vrm,na.rm=T),upt=mean(upt,na.rm=T),
              area=mean(area,na.rm=T)) %>%
          mutate(vrm.per.pop=vrm/pop.sa,vrm.per.area=vrm/area,Year=year(Date)) %>% ungroup() 



gg1<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>%  group_by(Year,Modes) %>% 
    summarize(vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T)) %>%
    ggplot(aes(x=Year,y=(vrm/10^6),color=Modes)) + geom_line() + 
    ggtitle("Annual VRM at Met Transit by Mode") + scale_y_continuous("Million VRM",labels =                                                            scales::comma,breaks=seq(0,40,1)) +
    scale_x_continuous(breaks=seq(2002,2019,2))
gg1

gg2<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>%  group_by(Year,Modes) %>% 
    summarize(vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T)) %>%
    ggplot(aes(x=Year,y=(upt/10^6),color=Modes)) + geom_line() + 
    ggtitle("Annual UPT at Met Transit by Mode") + scale_y_continuous("Million UPT",labels =                                                            scales::comma,breaks=seq(0,80,5)) +
    scale_x_continuous(breaks=seq(2002,2019,2))
gg2

#Percent UPT per vrm
gg3<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>%  group_by(Year,Modes) %>% 
    summarize(vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T)) %>%
    ggplot(aes(x=Year,y=(upt/10^6)/(vrm/10^6),color=Modes)) + geom_line() + 
    ggtitle("Annual UPT/VRM at Met Transit by Mode") + 
    scale_y_continuous("Million UPT/VRM",labels =                                                            scales::comma,breaks=seq(0,10,1)) +
    scale_x_continuous(breaks=seq(2002,2019,2))
gg3

gg4<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>%  group_by(Year,Modes,MT) %>% 
    summarize(vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T)) %>%
    mutate(gp=paste0(MT,"-",Modes)) %>%
    ggplot(aes(x=Year,y=(upt/10^6)/(vrm/10^6),color=gp)) + geom_line() + 
    ggtitle("Annual UPT/VRM at Met Transit by Mode broken down") + 
    scale_y_continuous("Million UPT/VRM",labels =                                                            scales::comma,breaks=seq(0,10,1)) +
    scale_x_continuous(breaks=seq(2002,2019,2))
gg4


gg5<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>%  group_by(Year,Modes,MT) %>% 
    summarize(vrm=sum(vrm,na.rm=T),upt=sum(upt,na.rm=T)) %>%
    mutate(gp=paste0(MT,"-",Modes)) %>%
    ggplot(aes(x=Year,y=(upt/10^6),color=gp)) + geom_line() + 
    ggtitle("Annual UPT by Mode broken down") + 
    scale_y_continuous("Million UPT",labels =                                                            scales::comma,breaks=seq(0,80,2)) +
    scale_x_continuous(breaks=seq(2002,2019,2))
gg5

#monthly UPT over time for Met Transit 

gg6.m<- gg1.df %>% filter(MT %in% c("MB","METC","MT")) %>% 
    mutate(gp=paste0(MT,"-",Modes)) %>%
    ggplot(aes(x=Date,y=(upt/10^6),color=gp)) + geom_line() + 
    ggtitle("Monthly UPT by Mode broken down") + 
    scale_y_continuous("Million UPT",labels =                                                            scales::comma,breaks=seq(0,80,2)) +
   scale_x_date(date_labels = "%b %y",date_breaks = "2 months") + 
   theme(axis.text.x =element_text(angle=90))
gg6.m


#Significant monthly fluctation. Worth looking at.
monthly.fluc<-gg1.df %>% filter(Year>2015, Year<2019,MT %in% c("MB","METC","MT"))  %>%
      group_by(month(Date),MT,Modes) %>%
    summarize(vrm.m=mean(vrm,na.rm=T), 
              vrm.sum=sum(vrm,na.rm=T),
              upt.m=mean(upt,na.rm=T),
              vrm.sd=sd(vrm,na.rm=T),
              upt.sf=sd(upt,na.rm=T)) %>%
  ggplot(aes(x=`month(Date)`,y=vrm.sum/10^6,color=paste0(MT,"-",Modes))) + geom_line() +   
  scale_x_continuous(breaks=seq(1,12,1)) + scale_y_continuous("Mil. VRM",breaks=seq(0,7,1))
  ggtitle("Cum Mil VRM  by month (2015-2018)") 

monthly.fluc

monthly.fluc2<-gg1.df %>% filter(Year>2015, Year<2019,MT %in% c("MB","METC","MT"))  %>%
      group_by(month(Date),MT,Modes) %>%
    summarize(vrm.m=mean(vrm,na.rm=T), 
              vrm.sum=sum(vrm,na.rm=T),
              upt.m=mean(upt,na.rm=T),
              vrm.sd=sd(vrm,na.rm=T),
              upt.sf=sd(upt,na.rm=T)) %>%
  ggplot(aes(x=`month(Date)`,y=vrm.m/10^6,color=paste0(MT,"-",Modes))) + geom_line() +   
  scale_x_continuous(breaks=seq(1,12,1)) + scale_y_continuous("Mil. VRM",breaks=seq(0,7,1)) + 
  ggtitle("Mean VRM  by month (2015-2018)")   

monthly.fluc2
  
monthly.fluc3<-gg1.df %>% filter(Year>2015, Year<2019,MT %in% c("MB","METC","MT"))  %>%
      group_by(month(Date),MT,Modes) %>%
    summarize(vrm.m=mean(vrm,na.rm=T), 
              vrm.sum=sum(vrm,na.rm=T),
              upt.m=mean(upt,na.rm=T),
              vrm.sd=sd(vrm,na.rm=T),
              upt.sf=sd(upt,na.rm=T)) %>%
  ggplot(aes(x=`month(Date)`,y=upt.m/10^6,color=paste0(MT,"-",Modes))) + geom_line() +   
  scale_x_continuous(breaks=seq(1,12,1)) + scale_y_continuous("Mil. UPT",breaks=seq(0,7,1)) +
  ggtitle("Mean Upt by month (2015-2018)")   
  
monthly.fluc3  
#gg1 %>% ggplotly()



```















